---
title: "Sipsey floodplain 4th corner analysis"
output:
  html_document:
    df_print: paged
    toc: yes
    toc_float: yes
  html_notebook:
    toc: yes
    toc_float: yes
---
Questions to answer:<br />
1.)	Where are species found across the floodplain?<br />
  a.	Simple analysis of plots <br />
2.)	What are the drivers that are leading to this distribution?<br />
  a.	Manyglm to test for significant drivers<br />
  b.	Multivarient SDM (part of the mvabund package)<br />
3.)	What are the drivers of ecosystem functions in BLH forests (based on the tree species)? The functions I’m looking at are; nutrient uptake (Carbon, Nitrogen, Phosphorus), water storage, and wildlife food production.<br /> 
  a.	Pulled data from the Try trait database for the dominant trees in the forest.<br />
  b.	Ran fourth corner analysis <br />
See https://heather-grab.github.io/Entom-4940/rql.html for more information on each step\
#Step 1: Workspace orginization/data read in
```{r,warning=FALSE,error=TRUE,echo=FALSE,results=FALSE}

#Gather libraries of interest
library(tidyverse)
library(vegan)
library(readxl)
library(forestmangr)
library(dplyr)
library(readr)
library(ggplot2)
library(patchwork)
library(mvabund)
library(lattice)

setwd("C:\\Users\\Matthew\\Desktop\\sipsydata\\Sipsey_tree_data\\Sipsey_tree\\4corneranalysis\\")
#read in data
#species code for later
spec_code=read.csv("data/species_code.csv")
spec_code=spec_code %>% mutate(AccSpeciesName=scientific.name)


#load R matrix R: the site by environment matrix that describes the environmental conditions at each site.
envi <- read.csv("data/masterdata.csv", row.names = 1)
envi=envi[,c(1,2,3,4,13,14)]



#load the Q matrix the species by traits matrix that describes the traits of each species (pull from data base)
my_data <- read.csv("data//trait.csv")
my_data=left_join(my_data,spec_code)
q=my_data %>% 
  select(code,TraitName,OrigValueStr) %>%  
  group_by(code,TraitName) %>% 
  #pivot q wider
  pivot_wider(names_from = TraitName, values_from=OrigValueStr,values_fn=mean)
#write to CSV
write_csv(q,"data//traits.csv")
traits=read.csv("data/traits.csv")
#Filter out traits we have data for
traits=traits %>% 
  select(code, Species.tolerance.to.waterlogging,Species.habitat.characterisation...Plant.requirement..soil.pH,
         Leaf.nitrogen.phosphorus..N.P..ratio,Leaf.area.per.leaf.dry.mass..specific.leaf.area..SLA.or.1.LMA...petiole.excluded,
         Leaf.area.per.leaf.dry.mass..specific.leaf.area..SLA.or.1.LMA...undefined.if.petiole.is.in..or.excluded,Leaf.phosphorus..P..content.per.leaf.dry.mass,
         Leaf.nitrogen..N..content.per.leaf.dry.mass) %>% 
  na.omit() %>% 
  #clean up
  mutate(code=code,'waterlog tolerance'=Species.tolerance.to.waterlogging,soilpH=Species.habitat.characterisation...Plant.requirement..soil.pH,
         'NP Ratio g/cm3'=Leaf.nitrogen.phosphorus..N.P..ratio, 'LMA petiol undefined mm2 mg-1'=Leaf.area.per.leaf.dry.mass..specific.leaf.area..SLA.or.1.LMA...undefined.if.petiole.is.in..or.excluded,
         'LMA petiol excluded mm2 mg-1'=Leaf.area.per.leaf.dry.mass..specific.leaf.area..SLA.or.1.LMA...petiole.excluded,
         'leaf N mg/g'= Leaf.nitrogen..N..content.per.leaf.dry.mass, 'leaf P mg/g'= Leaf.phosphorus..P..content.per.leaf.dry.mass, 
        .keep="unused")
write_csv(traits,"data//traitsfinal.csv")
traits=read.csv("data/traitsfinal.csv", row.names = 1)
#load L matrix: the site by species matrix that describes which species occur at each site
abd <- read_csv("data/nmds.csv",col_names = T)
rownames(abd)<-c(0:49)
```
#Step 2: Explore the data/manyglm model creation<br />
The goal here is to explore some potential models to see what covarients have a significant effect on tree species distribution.
```{r,warning=FALSE,error=TRUE,message=FALSE,fig.show=TRUE}
#look at abundance across the floodplain
Tree_spp=mvabund(abd)
rownames(Tree_spp)<-c(0:49)
plot(Tree_spp)
plot(Tree_spp)


#manyglm Plot to see if there is a pattern/cone 
mod1 <- manyglm(Tree_spp ~ envi$Hand, family="poisson")
plot(mod1)
mod2= manyglm(Tree_spp ~ envi$Hand, family="negative_binomial")
plot(mod2)
mod3=manyglm(Tree_spp~envi$Hand+envi$`Dist.from.river`+envi$plotsd, family = "negative_binomial")
plot(mod3)


#hypothesis testing with Anova
anova(mod3, nBoot=99)
anova(mod2, p.uni="adjusted", nBoot=99)
```
Initial results are indicating that HAND and distance from river have significant effects on species distribution.

#Step 3 multivarient SDM (https://cran.r-project.org/web/packages/mvabund/mvabund.pdf)\
The simplest form is a multivariate species distribution model (SDM) which predicts the abundance of each species based only on\ the environmental variables. The form of this model is essentially a negative binomial GLM with species as a fixed effect and a\ species by envi interaction.
```{r,warning=FALSE,error=TRUE,message=FALSE}
sdm_fit=traitglm(abd,envi)
sdm_fit$fourth
#plot
a        = max( abs(sdm_fit$fourth.corner) )
colort   = colorRampPalette(c("blue","white","red")) 
plot.spp = levelplot(t(as.matrix(sdm_fit$fourth.corner)), xlab="Environmental Variables",
                     ylab="Species", col.regions=colort(100), at=seq(-a, a, length=100),
                     scales = list( x= list(rot = 45)))
print(plot.spp)
#Run ANOVa on the model
anova.traitglm(sdm_fit)
```
The heat map resulting from this data is a good visual representation of each environmental variables effect on specific\ tree species

#Step 4: Forth corner analysis\
The forth corner analysis functions similarly to the multivarient SDM but looks at the environmental varibles effect on tree\
traits. The trait data was downloaded from the TRY plant trait database and more data will soon be added from the NEON database.
```{r}

#forth corner analysis
traits=traits
abd4th=abd %>% 
  mutate(QUMI=QUMI.1) %>% 
  select(ACRU,CACA18,CACO15,DIVI5,LIST2,QULA3,QUNI,QUMI,ULAM) 
fit=traitglm(abd4th,envi,traits)
fit$fourth 
#print fourth corner terms
#plot residuals
plot(fit)
#Check anova
anova(fit, nBoot = 10)
#summary can give you the effects of indivudual traits by environment interactions
#this is slow use nBoot=10 for test but 1000 for final analysis
summary(fit, nBoot=10)
#create heatmap
a        = max( abs(fit$fourth.corner) )
colort   = colorRampPalette(c("blue","white","red")) 
plot.4th = levelplot(t(as.matrix(fit$fourth.corner)), xlab="Environmental Variables",
                     ylab="Species traits", col.regions=colort(100), at=seq(-a, a, length=100),
                     scales = list( x= list(rot = 45)))
print(plot.4th)

#ANOVA
anova.traitglm(fit)
```
The relationships seen in the heat map appear promising. Next steps for me are to:\
•	Check NEON site for additional data/data we were missing from the Try database. \
•	Find papers that have paired the physiological traits to a quantifiable function.   


Note: As I work through the data analysis I'll update this document 
